### 2015-07-29 background.tex
fixes
update towards camera ready and also included the reviews

Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux driver. The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} struct as a parameter. However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order. Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break; // racy
    default: offset = -1;
  }
  if (offset < 0) return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\vspace{-2mm}
\caption{Racy entry point in the generic\_nvram Linux driver}
\label{fig:data_race_example}
\vspace{-2mm}
\end{figure}

@@

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break;
    default: offset = -1;
  }
  if (offset < 0) {
    mutex_unlock(&nvram_mutex); // unlock
    return -EINVAL;
  }
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\vspace{-2mm}
\caption{Using a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
\vspace{-2mm}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
\vspace{-2mm}
\caption{Applying lockset analysis on a concurrent program}
\label{fig:locksets}
\vspace{-2mm}
\end{figure}

>>>>

Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver. The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} struct as a parameter. However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order. Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break; // racy
    ...
  }
  if (offset < 0) return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\vspace{-2mm}
\caption{Racy entry point in a Linux 4.0 driver (drivers/char/generic\_nvram.c)}
\label{fig:data_race_example}
\vspace{-2mm}
\end{figure}

@@

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break;
    ...
  }
  if (offset < 0) {
    mutex_unlock(&nvram_mutex); // unlock
    return -EINVAL;
  }
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\vspace{-2mm}
\caption{Using a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
\vspace{-2mm}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
\vspace{-5mm}
\caption{Applying lockset analysis on a concurrent program}
\label{fig:locksets}
\vspace{-3mm}
\end{figure}

### 2015-07-29 corral.tex
fixes
update towards camera ready and also included the reviews

\noindent\textbf{Sound Partial-Order Reduction }
%
By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound. Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, then we do not instrument a \texttt{yield} after this statement.  This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context-switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.  This is because each access, for which a \texttt{yield} is suppressed, is guaranteed to be protected by some lock (a consequence of lockset analysis).  If the access is a write, its effects are not visible by the other entry point in the pair until the lock is released.  If the access is a read, the value of the shared location cannot change until the lock is released.  The fact that a \texttt{yield} is placed after each unlock operation suffices to take account of communication between entry points via the shared memory location.

@@

Our partial-order reduction can be used to accelerate \corral for arbitrary bug-finding in concurrent programs. In the remainder of this paper, though, we regard races as bugs and use \corral solely to look for races.

>>>>

\noindent\textbf{Sound Partial-Order Reduction }
%
By default, and assuming no race-freedom guarantees, \whoop instruments a \texttt{yield} after each shared memory access of each entry point, and after every lock and unlock operation.\footnote{We acknowledge that in the presence of data races and relaxed memory, even considering all interleavings of shared memory accesses may be insufficient to find all bugs.} \whoop then sends this instrumented program to \corral, which leverages sequentialization to explore all possible thread interleavings up to a pre-defined bound. Our approach to accelerating \corral is simple and yet effective: if, thanks to \whoop's analysis, we know that a given statement that accesses shared memory cannot be involved in a data race, then we do not instrument a \texttt{yield} after this statement.  This is a form of \emph{partial order reduction}~\cite{DBLP:books/sp/Godefroid96}, and reduces the number of context-switches that \corral must consider in a \emph{sound} manner: there is no impact on the bugs that will be detected.  This is because each access for which a \texttt{yield} is suppressed is guaranteed to be protected by some lock (a consequence of lockset analysis).  If the access is a write, its effects are not visible by the other entry point in the pair until the lock is released.  If the access is a read, the value of the shared location cannot change until the lock is released.  The fact that a \texttt{yield} is placed after each unlock operation suffices to take account of communication between entry points via the shared memory location.

@@

Our partial-order reduction can be used to accelerate \corral for arbitrary bug-finding in concurrent programs. In the remainder of this paper, though, we only regard races as bugs and use \corral solely to look for races.

### 2015-07-29 evaluation.tex
update towards camera ready and also included the reviews

\noindent\textbf{Race-Checking }
%
Table~\ref{tab:stats} presents statistics for all our benchmarks: lines of code (LoC); number of entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found). Using a higher csb did not uncover any further races; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug. \corral did not discover any races that \whoop did not already report.  This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.

@@

\whoop uses over-approximation to achieve scalability and, as expected, executes significantly faster than \corral in all our benchmarks. For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds. We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might get \corral stuck. This is not an issue for \whoop, which uses procedure summarization. This shows that \whoop has value as a stand-alone analyzer.

Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr. Figure~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, using a csb of 9. We noticed that a higher csb typically results into greater speedups when exploiting \whoop. This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
%
However, if \whoop fails to verify any race-related properties of a driver, it might slow down \corral. We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but it did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver quite quickly (21.4 seconds with csb of 9). We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.

\noindent\textbf{Other Tools }
%
We tried to compare \whoop with other similar approaches (see \S\ref{related}). However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007 and might be missing some of the latest features); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.

>>>>

\noindent\textbf{Race-Checking }
%
Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found). Using a higher csb did not uncover any further races; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug. \corral did not discover any races that \whoop did not already report.  This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.

@@

\whoop uses over-approximation to scale and, as expected, executes significantly faster than \corral in all our benchmarks. For example, \corral times out in all configurations (with and without \whoop) when trying to analyze the r8169 ethernet driver, while \whoop manages to analyze the driver in 144.5 seconds. We believe that the reason behind this is that r8169 has deeply-nested recursion in some of its entry points, which might cause \corral to get stuck. This is not an issue for \whoop, which uses procedure summarization. This shows that \whoop has value as a stand-alone analyzer.

Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr. Figure~\ref{fig:plot} shows that most speedups using \yieldmr are between 1.5$\times$ and 10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, with a csb of 9. We noticed that a higher csb typically results into greater speedups when exploiting \whoop. This is expected, as a higher csb translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce this sequentialization.
%
However, there are some cases where \whoop might slow down \corral. We noticed this in the sx8 driver: \whoop verified 46 out of 47 memory regions, but did not fully verify any of the pairs; \corral, on the other hand, analyzed the only two pairs of the driver in just 21.4 seconds (csb of 9). We believe that in this case the overhead of running \whoop outweighed the benefits of using \yieldmr.

\noindent\textbf{Other Tools }
%
We tried to compare \whoop with other similar approaches (see \S\ref{related}). However, we found this to be hard in practice: we downloaded Locksmith~\cite{pratikakis2006locksmith}, but could not get it to work with the 4.0 Linux drivers (the tool was last updated in 2007); we also could not find source code or binaries for \cite{kahlon2007fast, kahlon2009semantic, das2015section}.

### 2015-07-29 related.tex
update towards camera ready and also included the reviews

Static race analysis is a promising alternative to dynamic techniques, which restrict analysis to the schedule chosen by a (possibly controlled) scheduler, providing limited coverage~\cite{musuvathi2008finding}. Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable static race analyzers.  In comparison to \whoop, these tools rely heavily on user annotations.
%
Most related to our lockset analysis are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}. Both tools, though, have several limitations. RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races. To tackle this issue, RELAY employs unsound post-analysis filters and, hence, can also filter out real bugs. Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, showcasing its limited scalability. \whoop aims to achieve scalability \emph{and} precision: the first, via novel symbolic pairwise lockset analysis; and the second, by accelerating \corral, an industrial-strength precise bug-finder.

Prior works on static race checking also include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speedup dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve a scalable race analysis; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection. Our tool, \whoop, uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then uses \corral to filter out false races.

>>>>

Static race analysis is a promising alternative to dynamic techniques, which restrict analysis to the schedule chosen by a (possibly controlled) scheduler, providing limited coverage~\cite{musuvathi2008finding}. Warlock~\cite{sterling1993warlock} and LockLint~\cite{oracle2010locklint} are notable static race analyzers.  In comparison to \whoop, these tools rely heavily on user annotations.
%
Most related to our lockset analysis are the static lockset analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}. Both tools, though, have several limitations. RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races. To tackle this issue, RELAY employs unsound post-analysis filters and, hence, can also filter out real bugs. Locksmith was successfully applied in several small Pthreads applications and 7 medium-sized Linux device drivers, but the authors reported that the tool was unable to run on several large programs, hinting at its limited scalability. \whoop aims to achieve scalability \emph{and} precision: the first, via novel symbolic pairwise lockset analysis; and the second, by accelerating \corral, an industrial-strength precise bug-finder.

Prior works on static race checking also include: \cite{choi2002efficient}, which combines static analysis and runtime access caching to speed up dynamic race detection; \cite{kahlon2007fast}, which uses a divide-and-conquer algorithm that partitions all pointers of a program that do not alias in disjoint sets to achieve scalability; \cite{kahlon2009semantic}, which uses abstract interpretation to achieve a sound partial-order reduction on the set of thread interleavings and statically reduce the number of false race warnings; and \cite{das2015section}, which employs inter-procedural alias analysis and verifiable user annotations to split programs into disjoint sections, based on non-communicating accesses of shared data, and eliminate redundant checks during dynamic race detection. Our tool, \whoop, uses symbolic lockset analysis, which involves generating verification conditions and discharging them to a theorem prover, and then uses \corral to filter out false races.

### 2015-07-29 whoop.tex
update towards camera ready and also included the reviews

Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location. We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.  Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself. However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.  In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct (i.e.\ the assertion, which is --- described above --- at the end of the program holds), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

@@

\noindent\textbf{Extracting Entry Point Information }
%
Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions. Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.

@@

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentations and optimizations.
%
Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. However, it is relatively easy to enhance our tool with knowledge of other locking primitives. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. There are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

@@

Statically analyzing drivers requires ``closing'' the environment by abstracting away the low-level implementation details. To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in false positives. However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the underlying kernel functionality, without losing too much precision. Making our model more precise is an ongoing manual effort, but requires Linux expertise. We argue that further work on the model is orthogonal to the contributions of this paper. Also, even if our symbolic analysis results in false positives, \whoop can still use the results to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.

>>>>

Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location. We soundly model the effects of additional threads by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.  Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.
\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself. However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}
The trade-off is that a quadratic number of entry point pairs must be checked.  In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

@@

\noindent\textbf{Extracting Entry Point Information }
%
Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel functions. Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for \texttt{file\_operations}); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.

@@

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentations and optimizations.
%
Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. There are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

@@

Statically analyzing drivers requires ``closing'' the environment by abstracting away the low-level implementation details. To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in false positives. However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the Linux kernel, without losing too much precision. Making our model more precise is an ongoing manual effort, but requires Linux kernel expertise. We argue that further work on the model is orthogonal to the contributions of this paper. Also, even if our symbolic analysis results in false positives, \whoop can still use the results to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.
